Synthesizing Quotient Lenses

### 2018-06-29 papers/QuotientLenses/main.tex

Recent work~\cite{optician} developed a tool called Optician to synthesize a
subset of the {\em bijective lenses} that can be derived in the lens
programming language {\em Boomerang}~\cite{boomerang}. Each lens $\ell : S
\Leftrightarrow T$ that Optician can synthesize specifies a bijection from the
language of a source regular expression $S$ to a target language of a target
regular expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses easier.  However, many useful lenses
are not strictly bijective in nature---the desired transformation might ignore
whitespace or the exact ordering of data fields, for instance---and it is a
shame that Optician's synthesis algorithm cannot be directly applied in such
cases.  One observation, however, is that non-bijective transformations can
often be structured as a bijective ``core'' surrounded by some kind of data
normalization at the edges. We can therefore hope to use the Optician algorithm
as component in a system that synthesizes more complex lenses.

@@

Optician synthesizes bijective lenses from a pair $(S, T)$ of regular
expressions
specifying the source and target types and a set of example input-output pairs
that guide the synthesis algorithm. This presents a challenge for synthesizing
quotient lenses since a specification of its source and target formats
needs to account for the equivalence relations defined on the
respective formats. Our solution is to introduce {\em Quotient Regular
Expressions} (QREs), which are regular expressions augmented with extra
syntax that enables programmers to simultaneously specify a regular expression
and an equivalence relation on the language of that regular expression.
Further, given a QRE, we can automatically infer a \emph{canonizer}---a
function that converts strings in the language of the regular expression
to a canonical form.

@@

% Our end goal is to synthesize bijective quotient lenses that map between QREs
% via a bijective lens between the respective canonical formats of the QREs. This
% idea is simple and natural, but begs the following question: do we give up
% expressiveness if we restrict ourselves to this form? We will show that this
% is not the case by doing the following. First, we will introduce {\em QRE
% lenses}, which are bijective quotient lenses formed by ``lifting'' bijective
% lenses to a bijective quotient lens type, then freely quotienting these lenses
% by QREs or applying familiar regular and composition operators to them. Then,
% we will prove a normal form theorem (Theorem~\ref{normal form})
% that says that every QRE lens can be transformed into a functionally equivalent
% lens that canonizes source and target data only once at the ``edges'' then uses
% a bijective lens to map between the respective canonical elements; no internal
% canonization occurs in a lens in this normal form. This normalization
% property will enable us to (1) synthesize QRE lenses by extending the synthesis
% algorithm used by Optician, and (2) prove that if there is a QRE lens that
% satisfies the input specification, then this extended algorithm will return
% such a lens.

@@

Recent work described the Optician
algorithm/tool~\cite{optician}, which can synthesize bijective lenses
such as \cd{bib_to_end}, obviating the sometimes tedious tasks
involved in writing such lenses by hand.  Given the directive
\begin{lstlisting}
synth S $\Leftrightarrow$ T using exs
\end{lstlisting}
\noindent
Optician will synthesize a bijective lens between 
source and target formats described by regular expressions
\cd{S} and \cd{T}, respectively, and constrained by a set of input--output example pairs \cd{exs} specifying how
the synthesized lens should behave on those examples.

@@

Recall that given regular expressions $R, S$ and equivalence relations
$\equiv_R$ and $\equiv_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$, a
{\em bijective quotient lens} $q : R /{\equiv_R} \Leftrightarrow S/{\equiv_S}$
is a pair of functions $q.\get : \mathcal{L}(R) \longrightarrow \mathcal{L}(S)$
and $q.\lput : \mathcal{L}(S) \longrightarrow \mathcal{L}(R)$ such that for all
$r \in \mathcal{L}(R)$ and $s \in \mathcal{L}(S)$, we have $q.\lput(q_1.\get(r))
\equiv_R r$ and $q.\get(q_1.\lput(s)) \equiv_R s$. Moreover, if $r \equiv_R
r'$, then $q.\get(r) = q.\get(r')$, and if $s \equiv_S s'$, then $q.\lput(s) =
q.\lput(s')$. In words, a bijective quotient lens $q$ from $R$ to $S$ modulo
$\equiv_R$ and $\equiv_S$ is a pair of functions $q.\get$ and $q.\lput$ such
that $q.\get$ respects $\equiv_R$ and $q.\lput$ respects $\equiv_S$, and such
that the {\em get} and {\em put} functions lifted to $\mathcal{L}(R)/{\equiv_R}$
and $\mathcal{L}(S)/{\equiv_S}$ are mutual inverses. These laws are similar
bijective lens laws \cref{bijectivelenslaws}, except that the equality
restrictions in the bijective lens laws are loosened to allow for equivalence
relations. Also, the condition that $r \equiv_R r'$ implies $q.\get(r) =
q.\get(r')$ ensures that the \kw{get} function induced on the equivalence
classes of $\equiv_R$ is well-defined, and similarly for the \kw{put}
function.

@@

Each bijective string lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where
$R$ and $S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the
source language of $\ell$ is $\mathcal{L}(R)$ and the target language of $\ell$
is $\mathcal{L}(S)$. IThe final rule in the figure deserves special attention.
It states that a lens $\ell : R \Leftrightarrow S$ can also be considered to be
of type $\ell : R' \Leftrightarrow S'$ provided that $R$ (resp. $S$) can be
proven to be equivalent to $R'$ (resp. $S$) from the star-semiring axioms
(associativity and commutativity of | and $\cdot$, identity of the empty string
$\epsilon$, distibutivity of $\cdot$ over |, annihilative property of
$\varnothing$ with respect to $\cdot$, and that $R^* = \epsilon \; | \; R \cdot
R^* = \epsilon \; | \; R^* \cdot R$ for all $R$):

@@

The Optician framework~\cite{optician} showed how to do such lens synthesis in
the case for bijective lenses. Here we show how to reduce QRE lens synthesis to
that case, so that we can re-use the Optician algorithm but in the more
expressive context of QRE lenses.  The basic idea is straightforward: we run the
Optician algorithm to synthesize a lens between the kernels of two QREs and then
apply the canonizers at the edges to recover a lens between the whole
languages.  This simple strategy turns out to be remarkably effective, and the
idea of using Optician in this way inspired the design of our QRE lenses.

The Optician bijective lens synthesis algorithm works as follows. Given regular
expressions $R$ and $S$ and a set of example input--output pairs $\{(r_1,
s_1),\allowbreak \ldots,\allowbreak (r_n, s_n)\}$, Optician will try to find a
bijective lens $\ell : R \Leftrightarrow S$ that agrees on the examples,
\textit{i.e.}, it maps $r_i$ to $s_i$ in the forward direction and $s_i$ to
$r_i$ in the backward direction. The bijective lens synthesis algorithm is
guaranteed to succeed (eventually!) if such a bijective lens exists. There are
typically a large number of such lenses, so this algorithm chooses the one that
corresponds to a minimal \emph{alignment} of the regular expressions. Additional
details on this algorithm can be found in the original paper on synthesizing
bijective lenses~\cite{optician}.

@@

%
\begin{itemize}
  \item[\QRESize{}:] 
  The number of AST nodes in the QRE specifications for the source and
  target formats, including examples. 
  \item[\canonizeAndSpecSize{}:] 
  The sum of (1) the number of AST nodes in $W(q)$ for each QRE $q$ in the source and target
  formats, (2) the number of AST nodes in $\canonize(q)$ for each QRE $q$ with a
  non-trivial canonizer, and (3) the number of AST nodes in the
  examples.  We use (1) to estimate the burden of describing
  the external source and target formats and (2) to estimate the
  burdern of writing the requisite canonizers
  by hand.  We count the nodes in the examples because they would be
  fed to the bijective synthesizer.  
  These counts are an approximation, as both $W(q)$ and $\canonize(q)$ are
  automatically generated from the corresponding QRE $q$, and it is
  possible that a human-written version might be smaller.
  \item[\LensAndSpecSize{}:] The sum of (1) the number of AST nodes in
  $W(q)$ for each QRE $q$ in the source and target formats and (2) the
  number of AST nodes in the synthesized QRE lens.  We use (1) to
  estimate the burdern of describing the source and target formats
  and (2) to estimate the burdern of writing the appropriate lens by
  hand. These counts are also approximations, as
  $W(q)$ and the synthesized lens may be larger than one written by hand.
\end{itemize}

@@

A newer language which supports quotenting is BiFluX~\cite{pacheco2014biflux}, a
bidirectional functional update language for XML. BiFluX is inspired by
the FLUX XML update language~\cite{cheney2008flux}, and adopts a bidirectional
programming by update paradigm, where a program succintly and precisely
describes {\em how} to update a source document with a target document, in an
intuitive way, such that there is a unique ``inverse'' source query for each
update program. The source and view types of BiFluX programs are given by
regular expressions.

In BiFluX, a regular expression $R$ is said to be unambiguous if there is only
one way to parse a {\em flat} value of type $R$ to a stuctured value of type
$R$; here a flat value of $R$ is a value of $R$ with all left/right tags,
parentheses and list brackets removed. The bidirectional part of BiFluX involves
transforming source regular expressions to view regular expressions so that
data sequences described by the two types can be matched. This process can
result in intermediate types that are ambiguous, and this ambiguity can cause
unintended updates to be made the source, especially when information is
discarded in this process.

\iffalse
For example, imagine that a book database has source type \\
$S$ = \lstinline{books[book[title[string], author[string]+]*]}
with view type $V$ = \lstinline{title[string]}. When the source type is
transformed to the view type for matching, only titles are kept, the element
labels are destroyed and all authors are replaced for the empty sequence, thus
leaving the intermediate type $S'$ = \lstinline{(title[string],(),()*)*}. In
BiFluX, $V$ is considered a subtype of $S'$ because the flat values of $V$ are
also flat values of $S'$. Since the intermediate type is $S'$ ambiguous, a view
sequence \lstinline{[title["mybook"]]} would be upcast into an intermediate
value \lstinline{[(title["mybook"],((), [ ]))]} that would in turn lead to an
updated source \lstinline{books[[book[(title["mybook"],(aut, [ ]))]]]}, which is
clearly unsatisfactory as it would discard the entire book database except the
first author \lstinline{aut} of the first book.
\fi
The typechecking phase of a BiFluX program therefore includes a {\em type
normalization} phase that tries to normalize a source types using automata
reduction techniques, and derive a lens between ambiguous and unambiguous
types. On the view side, this phase only tries to normalize view types into
isomorphic types. Pacheco et. al. do not claim that their normalization
procedures are complete in the sense that they can disambiguate any ambiguous
type, but they do show that when these procedures succeed the normalized types
are unambiguous.

{\em Formlenses}~\cite{rajkumar2014lenses} are a variant of lenses that
perform transformations up to equivalence. Formlenses are a bidirectional
generalization of {\em formlets}~\cite{cooper2008essence}, which are a
high-level abstraction for building Web forms. Formlets encapsulate several
low-level details including selecting field names for elements and parsing data
from client responses. In their work, Rajkumar et. al. represent formlets as
functions that take a source of names---concretely, an integer $n$ that should
be used as the next name---as an argument and produce a triple consisting of an
HTML document, a {\em collect} function, and a modified namesource:
\begin{center}
\begin{lstlisting}
type Formlet a = Int $\to$ (Html, Env $\to$ a, Int)
\end{lstlisting}
\end{center}
The names of any generated form fields are drawn from the namesource, and the
{\em collect} function looks up precisely these names. For example the
\kw{inputInt} combinator below builds a form that accepts an integer,
\begin{center}
\begin{lstlisting}
inputInt :: Formlet Int
inputInt i = let n = show i in
             ($\kw{inputTag}$ {name = n, value = ""},$\lambda e \longrightarrow \kw{read}$ ($\kw{fromJust}$ ($\kw{lookup}$ n e)),i+1)
\end{lstlisting}
\end{center}
where \kw{inputTag} constructs an input element, \kw{lookup}
retrieves a value from an association list \kw{fromJust} extracts the
encapsulated value from a \kw{Maybe} value, and \kw{read} parses an \kw{Int}
from a \kw{String}.

Rajkumar et. al. observe that formlets are a useful abstraction, but they only
address one half of the problem---they make it easy to construct a function
that {\em produces} a value of type \codefont{a}, but they do not provide a way
to describe a function that {\em consumes} a value of type \codefont{a} and
embeds it in a form. A different way to think about this problem is to observe
that forms are often used to present an {\em updateable view} of the data
source. Therefore, Rajkumar et. al. leverage the power of bidirectional
transformations to define formlets that behave like updatable views, yielding
the abstraction of formlenses.

The Formlens type is given by
\begin{center}
\begin{lstlisting}
type Formlens a = Maybe a $\to$ [Int] $\to$ (Html,Env $\to$ Maybe a,[Int])
\end{lstlisting}
\end{center}
with the differences between this definition and the standard definition for
formlets being an extra \codefont{Maybe\; a} parameter that provides an
optional initial value of the form, the namesource being changed to a list of
integers, and the {\em collect} function being allowed to return an optional
value.

Rajkumar et. al. consider only formlenses that are well-formed in that a
well-formed formlens should only draw names from the namesource provided as an
argument, and the {\em collect} function should only look up corresponding
names. Under this assumption, Rajkumar et. al. define what it means for a
formlens to be well behaved by defining a version of the classical
\textsc{GetPut} and \textsc{PutGet} laws for formlenses. These laws hold modulo
an equivalence relation on environments, since a formlens may be invoked with
two separate namespaces, but this should not affect the value stored in the
information encapsulated in the type $a$. Consequently, the \textsc{GetPut} and
\textsc{PutGet} laws defined for formlenses are in spirit analogous to the
\textsc{GetPut} and \textsc{PutGet} laws for quotient lenses, since they hold
up to renaming of environment variables.

@@

Next we turn to related work in program synthesis. Much of the research in
synthesis assumes that the synthesizer is provided with a collection of examples.
Optician and Optician$_Q$ differ in that they require the programmer to supply
both examples {\em and} format descriptions in the form of regular expressions
or QREs, though these two systems are far from the only ones to consider
type-based synthesis. There is a trade-off here.  On the one hand, a user must
have some programming expertise to write regular expression (or QRE)
specifications, and it requires some work. On the other hand, such
specifications provide a great deal of information to the synthesis system,
which decreases the number of examples needed (often to zero), makes the system
scale well, and allows it to handle large, complex formats.  By providing these
format specifications, the synthesis engine does not have to both infer the
format of the data as well as the transformations on it, obviating the need to
infer tricky formats like those involving nested iterations.

There are many other recent results showing how to synthesize functions from
type-based
specifications~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,scherer-icfp-2015,frankle+:popl16,armando+:pldi16}.
These systems enumerate programs of their target language, orienting their
search procedures to process only terms that are well-typed.
Optician is distinctive in that it synthesizes terms in a language with many
type equivalences. Perhaps the system most similar to Optician is
InSynth~\cite{gvero-pldi-2013}, a system for synthesizing terms in the
simply-typed lambda calculus that addresses equivalences on types. Instead of
trying to directly synthesize terms of the simply-typed lambda calculus,
InSynth synthesizes a well-typed term in the succinct calculus, a language with
types that are equivalent ``modulo isomorphisms of products and
currying''~\cite{gvero-pldi-2013}. The type structure used in Optician
is significantly more complex.  In particular, because Optician types do not
have full canonical forms, Anders et al. used a pseudo-canonical form that
captures part of the equivalence relation over types were used. To preserve
completeness, they pushed some of the remaining parts of the type equivalence
relation into a set of rewriting rules and other parts into the synthesis
algorithm itself.

>>>>

Recent work~\cite{miltner2017synthesizing} developed a tool called Optician to synthesize a
subset of the {\em bijective lenses} that can be derived in the lens
programming language {\em Boomerang}~\cite{boomerang}. Each lens $\ell : S
\Leftrightarrow T$ that Optician can synthesize specifies a bijection from the
language of a source regular expression $S$ to a target language of a target
regular expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses easier.  However, many useful lenses
are not strictly bijective in nature---the desired transformation might ignore
whitespace or the exact ordering of data fields, for instance---and it is a
shame that Optician's synthesis algorithm cannot be directly applied in such
cases.  One observation, however, is that non-bijective transformations can
often be structured as a bijective ``core'' surrounded by some kind of data
normalization at the edges. We can therefore hope to use the Optician algorithm
as component in a system that synthesizes more complex lenses.

@@

Optician synthesizes bijective lenses from a pair $(S, T)$ of regular
expressions
specifying the source and target types and a set of example input-output pairs
that guide the synthesis algorithm. This presents a challenge for synthesizing
quotient lenses since a specification of a lens's source and target formats
needs to account for the equivalence relations defined on the
respective formats. Our solution is to introduce {\em Quotient Regular
Expressions} (QREs), which are regular expressions augmented with extra
syntax that enables programmers to simultaneously specify a regular expression
and an equivalence relation on the language of that regular expression.
Further, given a QRE, we can automatically infer a \emph{canonizer}---a
function that converts strings in the language of the regular expression
to a canonical form.

@@

% Our end goal is to synthesize bijective quotient lenses that map between QREs
% via a bijective lens between the respective canonical formats of the QREs. This
% idea is simple and natural, but provokes the following question: do we give up
% expressiveness if we restrict ourselves to this form? We will show that this
% is not the case by doing the following. First, we will introduce {\em QRE
% lenses}, which are bijective quotient lenses formed by ``lifting'' bijective
% lenses to a bijective quotient lens type, then freely quotienting these lenses
% by QREs or applying familiar regular and composition operators to them. Then,
% we will prove a normal form theorem (Theorem~\ref{normal form})
% that says that every QRE lens can be transformed into a functionally equivalent
% lens that canonizes source and target data only once at the ``edges'' then uses
% a bijective lens to map between the respective canonical elements; no internal
% canonization occurs in a lens in this normal form. This normalization
% property will enable us to (1) synthesize QRE lenses by extending the synthesis
% algorithm used by Optician, and (2) prove that if there is a QRE lens that
% satisfies the input specification, then this extended algorithm will return
% such a lens.

@@

Recent work described the Optician
algorithm/tool~\cite{miltner2017synthesizing}, which can synthesize bijective lenses
such as \cd{bib_to_end}, obviating the sometimes tedious tasks
involved in writing such lenses by hand.  Given the directive
\begin{lstlisting}
synth S $\Leftrightarrow$ T using exs
\end{lstlisting}
\noindent
Optician will synthesize a bijective lens between 
source and target formats described by regular expressions
\cd{S} and \cd{T}, respectively, and constrained by a set of input--output example pairs \cd{exs} specifying how
the synthesized lens should behave on those examples.

@@

Recall that given regular expressions $R, S$ and equivalence relations
$\equiv_R$ and $\equiv_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$, a
{\em bijective quotient lens} $q : R /{\equiv_R} \Leftrightarrow S/{\equiv_S}$
is a pair of functions $q.\get : \mathcal{L}(R) \longrightarrow \mathcal{L}(S)$
and $q.\lput : \mathcal{L}(S) \longrightarrow \mathcal{L}(R)$ such that for all
$r \in \mathcal{L}(R)$ and $s \in \mathcal{L}(S)$, we have $q.\lput(q.\get(r))
\equiv_R r$ and $q.\get(q.\lput(s)) \equiv_S s$. Moreover, if $r \equiv_R
r'$, then $q.\get(r) = q.\get(r')$, and if $s \equiv_S s'$, then $q.\lput(s) =
q.\lput(s')$. In words, a bijective quotient lens $q$ from $R$ to $S$ modulo
$\equiv_R$ and $\equiv_S$ is a pair of functions $q.\get$ and $q.\lput$ such
that $q.\get$ respects $\equiv_R$ and $q.\lput$ respects $\equiv_S$, and such
that the {\em get} and {\em put} functions lifted to $\mathcal{L}(R)/{\equiv_R}$
and $\mathcal{L}(S)/{\equiv_S}$ are mutual inverses. These laws are similar to the
bijective lens laws \cref{bijectivelenslaws}, except that the equality
restrictions in the bijective lens laws are loosened to allow for equivalence
relations. Also, the condition that $r \equiv_R r'$ implies $q.\get(r) =
q.\get(r')$ ensures that the \kw{get} function induced on the equivalence
classes of $\equiv_R$ is well-defined, and similarly for the \kw{put}
function.

@@

Each bijective string lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where
$R$ and $S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the
source language of $\ell$ is $\mathcal{L}(R)$ and the target language of $\ell$
is $\mathcal{L}(S)$. The final rule in the figure deserves special attention.
It states that a lens $\ell : R \Leftrightarrow S$ can also be considered to be
of type $\ell : R' \Leftrightarrow S'$ provided that $R$ (resp. $S$) can be
proven to be equivalent to $R'$ (resp. $S$) from the star-semiring axioms
(associativity and commutativity of | and $\cdot$, identity of the empty string
$\epsilon$, distibutivity of $\cdot$ over |, annihilative property of
$\varnothing$ with respect to $\cdot$, and that $R^* = \epsilon \; | \; R \cdot
R^* = \epsilon \; | \; R^* \cdot R$ for all $R$):

@@

The Optician framework~\cite{miltner2017synthesizing} showed how to do such lens synthesis in
the case for bijective lenses. Here we show how to reduce QRE lens synthesis to
that case, so that we can re-use the Optician algorithm but in the more
expressive context of QRE lenses.  The basic idea is straightforward: we run the
Optician algorithm to synthesize a lens between the kernels of two QREs and then
apply the canonizers at the edges to recover a lens between the whole
languages.  This simple strategy turns out to be remarkably effective, and the
idea of using Optician in this way inspired the design of our QRE lenses.

The Optician bijective lens synthesis algorithm works as follows. Given regular
expressions $R$ and $S$ and a set of example input--output pairs $\{(r_1,
s_1),\allowbreak \ldots,\allowbreak (r_n, s_n)\}$, Optician will try to find a
bijective lens $\ell : R \Leftrightarrow S$ that agrees on the examples,
\textit{i.e.}, it maps $r_i$ to $s_i$ in the forward direction and $s_i$ to
$r_i$ in the backward direction. The bijective lens synthesis algorithm is
guaranteed to succeed (eventually!) if such a bijective lens exists. There are
typically a large number of such lenses, so this algorithm chooses the one that
corresponds to a minimal \emph{alignment} of the regular expressions. Additional
details on this algorithm can be found in the original paper on synthesizing
bijective lenses~\cite{miltner2017synthesizing}.

@@

%
\begin{itemize}
  \item[\QRESize{}:] 
  The number of AST nodes in the QRE specifications for the source and
  target formats, including examples. 
  \item[\canonizeAndSpecSize{}:] 
  The sum of (1) the number of AST nodes in $W(q)$ for each QRE $q$ in the source and target
  formats, (2) the number of AST nodes in $\canonize(q)$ for each QRE $q$ with a
  non-trivial canonizer, and (3) the number of AST nodes in the
  examples.  We use (1) to estimate the burden of describing
  the external source and target formats and (2) to estimate the
  burden of writing the requisite canonizers
  by hand.  We count the nodes in the examples because they would be
  fed to the bijective synthesizer.  
  These counts are an approximation, as both $W(q)$ and $\canonize(q)$ are
  automatically generated from the corresponding QRE $q$, and it is
  possible that a human-written version might be smaller.
  \item[\LensAndSpecSize{}:] The sum of (1) the number of AST nodes in
  $W(q)$ for each QRE $q$ in the source and target formats and (2) the
  number of AST nodes in the synthesized QRE lens.  We use (1) to
  estimate the burdern of describing the source and target formats
  and (2) to estimate the burdern of writing the appropriate lens by
  hand. These counts are also approximations, as
  $W(q)$ and the synthesized lens may be larger than one written by hand.
\end{itemize}

@@

A newer language which supports quotenting is FliPpr \cite{matsuda2013flippr}. FliPpr is a program transformation system that uses program inversion to produce a CFG parser from a pretty-printer. FliPpr includes biased choice combinators that choose canonical representatives for sets of strings. For example, one can define variants of (white)spaces with the choice operator as,
\begin{center}
\begin{lstlisting}
nil = text "" <+ space
space = (text " " <+ text "\n") <> nil
\end{lstlisting}
\end{center}
where \lstinline|nil| and \lstinline|space| pretty-print \lstinline|""| and \lstinline|" "| respectively, but represent zero-or-more and one-or-more whitespaces in parsing. FliPpr also allows annotations for optional parentheses in programs. These features enable a user to easily define a pretty-printer that also recognizes ugly strings that get mapped to a canonical representative modulo whitespace and parentheses quotienting.

Another programming language that supports quotienting is BiFluX~\cite{pacheco2014biflux}, a
bidirectional functional update language for XML. BiFluX is inspired by
the FLUX XML update language~\cite{cheney2008flux}, and adopts a bidirectional
programming by update paradigm, where a program succintly and precisely
describes {\em how} to update a source document with a target document, in an
intuitive way, such that there is a unique ``inverse'' source query for each
update program. The source and view types of BiFluX programs are given by
regular expressions.

In BiFluX, a regular expression $R$ is said to be unambiguous if there is only
one way to parse a {\em flat} value of type $R$ to a structured value of type
$R$; here a flat value of $R$ is a value of $R$ with all left/right tags,
parentheses and list brackets removed. The bidirectional part of BiFluX involves
transforming source regular expressions to view regular expressions so that
data sequences described by the two types can be matched. This process can
result in intermediate types that are ambiguous, and this ambiguity can cause
unintended updates to be made the source, especially when information is
discarded in this process.

\iffalse
For example, imagine that a book database has source type \\
$S$ = \lstinline|books[book[title[string], author[string]+]*]|
with view type $V$ = \lstinline|title[string]|. When the source type is
transformed to the view type for matching, only titles are kept, the element
labels are destroyed and all authors are replaced for the empty sequence, thus
leaving the intermediate type $S'$ = \lstinline|(title[string],(),()*)*|. In
BiFluX, $V$ is considered a subtype of $S'$ because the flat values of $V$ are
also flat values of $S'$. Since the intermediate type is $S'$ ambiguous, a view
sequence \lstinline|[title["mybook"]]| would be upcast into an intermediate
value \lstinline|[(title["mybook"],((), [ ]))]| that would in turn lead to an
updated source \lstinline|books[[book[(title["mybook"],(aut, [ ]))]]]|, which is
clearly unsatisfactory as it would discard the entire book database except the
first author \lstinline|aut| of the first book.
\fi
The typechecking phase of a BiFluX program therefore includes a {\em type
normalization} phase that tries to normalize a source types using automata
reduction techniques, and derive a lens between ambiguous and unambiguous
types. On the view side, this phase only tries to normalize view types into
isomorphic types. Pacheco et. al. do not claim that their normalization
procedures are complete in the sense that they can disambiguate any ambiguous
type, but they do show that when these procedures succeed the normalized types
are unambiguous.

{\em Formlenses}~\cite{rajkumar2014lenses} are a variant of lenses that
perform transformations up to equivalence. Formlenses are a bidirectional
generalization of {\em formlets}~\cite{cooper2008essence}, which are a
high-level abstraction for building Web forms. Formlets encapsulate several
low-level details including selecting field names for elements and parsing data
from client responses. In their work, Rajkumar et. al. represent a formlens as a
function that takes an optional initial value of a form and a list of integers as a name source and produces a triple consisting of an
HTML document, a {\em collect} function, and a modified namesource:
\begin{center}
\begin{lstlisting}
type Formlens a = Maybe a $\to$ [Int] $\to$ (Html,Env $\to$ Maybe a,[Int])
\end{lstlisting}
\end{center}
The names of any generated form fields are drawn from the name source, and the
{\em collect} function looks up precisely these names

Rajkumar et. al. consider only formlenses that are well-formed in the sense that a
well-formed formlens should only draw names from the namesource provided as an
argument, and the {\em collect} function should only look up corresponding
names. Under this assumption, Rajkumar et. al. define what it means for a
formlens to be well behaved by defining a version of the classical
\textsc{GetPut} and \textsc{PutGet} laws for formlenses. These laws hold modulo
an equivalence relation on environments, since a formlens may be invoked with
two separate namespaces, but this should not affect the value stored in the
information encapsulated in the type $a$. Consequently, the \textsc{GetPut} and
\textsc{PutGet} laws defined for formlenses are in spirit analogous to the
\textsc{GetPut} and \textsc{PutGet} laws for quotient lenses, since they hold
up to renaming of environment variables.

@@

Next we turn to related work in program synthesis. Much of the research in
synthesis assumes that the synthesizer is provided with a collection of examples.
Optician and QRE-Enhanced Optician differ in that they require the programmer to supply
both examples {\em and} format descriptions in the form of regular expressions
or QREs, though these two systems are far from the only ones to consider
type-based synthesis. There is a trade-off here.  On the one hand, a user must
have some programming expertise to write regular expression (or QRE)
specifications, and it requires some work. On the other hand, such
specifications provide a great deal of information to the synthesis system,
which decreases the number of examples needed (often to zero), makes the system
scale well, and allows it to handle large, complex formats.  By providing these
format specifications, the synthesis engine does not have to both infer the
format of the data as well as the transformations on it, obviating the need to
infer tricky formats like those involving nested iterations.

There are many other recent results showing how to synthesize functions from
type-based
specifications~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,scherer-icfp-2015,frankle+:popl16,armando+:pldi16}.
These systems enumerate programs of their target language, orienting their
search procedures to process only terms that are well-typed.
Optician is distinctive in that it synthesizes terms in a language with many
type equivalences. Perhaps the system most similar to Optician is
InSynth~\cite{gvero-pldi-2013}, a system for synthesizing terms in the
simply-typed lambda calculus that addresses equivalences on types. Instead of
trying to directly synthesize terms of the simply-typed lambda calculus,
InSynth synthesizes a well-typed term in the succinct calculus, a language with
types that are equivalent ``modulo isomorphisms of products and
currying''~\cite{gvero-pldi-2013}. The type structure used in Optician
is significantly more complex.  In particular, because Optician types do not
have full canonical forms, Miltner et al. \cite{miltner2017synthesizing} used a pseudo-canonical form that
captures part of the equivalence relation over types. To preserve
completeness, they pushed some of the remaining parts of the type equivalence
relation into a set of rewriting rules and other parts into the synthesis
algorithm itself.
